---
title: "Machine Learning-Weight Lifting Exercises Dataset"
author: "John Shomaker"
date: "10/2/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Report Environment & Files

Scripts produced and executed on MAC OS X 10.11.6, and RStudio 0.99.903.

GitHub repo: https://github.com/jshomaker/ML-Fitness.git

### Background & Data

"Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways."

Training data:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

Test data:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

Citation:
http://groupware.les.inf.puc-rio.br/har

"The Weight Lifting Exercises Dataset (WLE) includes data for six (6) young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).""

### Objectives

Predict the manner ("classe" variable) in which participants did the exercise. Specific outputs include:

+ Demonstrate how model designed and built
+ Demonstrate cross-validation of training set (to reduce error or over-fit)
+ Demonstrate resulting out-of-sample error
+ Predict classe for 20 different test cases
+ Submit as a Github repo with Rmd/HTML (< 2000 wds, < 5 figures) (with gh-pages branch)


### 1. Load Relevant Machine Learning Libraries

``` {r load_libraries}

library(caret)
library(rpart)
library(randomForest)

set.seed(1226)

```

### 2. Download Training & Testing Datasets

```{r WLE_download}

## Download the training and testing sets for the fitness data

training_file <- "training.csv"
testing_file <- "testing.csv"

if (!file.exists(training_file)){
      training_URL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
      download.file(training_URL, training_file, method="curl")
}  

if (!file.exists(testing_file)){
      testing_URL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
      download.file(testing_URL, testing_file, method="curl")
}

training <- read.csv(training_file, na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(testing_file, na.strings=c("NA","#DIV/0!",""))

```

### 3. Subset Training Into Training & Validation Datasets

```{r cv_sets}

## Subset training into training (train1) (60%) and validation (val1) (40%)
## Val1 will provide out-of-sample error

inTrain <- createDataPartition(y = training$classe, p = 0.6, list = FALSE)
train1 <- training[inTrain,]
val1 <- training[-inTrain,]

train1$classe <- as.factor(train1$classe)
val1$classe <- as.factor(val1$classe)

```

### 4. Eliminate Predictors With Zeros, No Value Variation, and/or Mostly NAs

``` {r clean_data}

## Cleaning 1 - eliminate predictors with near-zero variance/variation

NZV <- nearZeroVar(train1, saveMetrics=TRUE)

## Note: found this code online to reduce typing

NZVnames <- names(train1) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt","kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt","max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm","var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm","stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
"kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
"max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
"kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell","skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell","amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm","skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm",
"max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm",
"amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm",
"avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm",
"stddev_yaw_forearm", "var_yaw_forearm")

train2 <- train1[!NZVnames]

## Clean 2 - eliminate row_ID, since not relevant predictor

train2 <- train2[c(-1)]

## Clean 3 - eliminate predictors with > 50% NAs

train3 <- train2

for(i in 1:length(train2)) { #for every column in the training dataset

      if( sum( is.na( train2[, i] ) ) /nrow(train2) >= .5 ) { #if NAs > 60% of total
        
             for(j in 1:length(train3)) {
                    
                    if( length( grep(names(train2[i]), names(train3)[j]) ) ==1)  { #if same
                
                           train3 <- train3[ , -j] # delete column
            }   
        } 
    }
}

## Clean 4 - apply final columns to val1 (validation dataset)

clean_col <- colnames(train3)
val3 <- val1[clean_col]

## Clean 5 - apply final columns without 'classe' to testing (official test dataset)

clean_col_wo_classe <- colnames(train3[, -58])
testing <- testing[clean_col_wo_classe]

dim(train3); dim(val3); dim(testing)

```

## 5. Compare Prediction & Error for Decision Tree vs. Random Forest

Selected Random Forest, since focused on prediction accuracy (albeit more difficult to interpret individual coefficients).

``` {r two_models}

## Random Forest running way too long with 10,000+ rows and 58 variables
## Found libraries to run CPU cores in parallel

library(parallel)
library(doParallel)
cluster <- makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)
fitControl <- trainControl(method = "cv", number = 10, allowParallel = TRUE)

## Two models, referenced as Decision Tree (DT) and Random Forest (RF)
## Train the models and then apply to val1 (test) for prediction error

modDT <- train(classe ~ ., method = "rpart", data = train3)
modRF <- train(classe ~ ., method="rf", data =  train3, trControl = fitControl)

## Release CPU core cluster
stopCluster(cluster)

predDT <- predict(modDT, val1)
predRF <- predict(modRF, val1)

confusionMatrix(predDT, val1$classe)
confusionMatrix(predRF, val1$classe)

```
Random Forest is much more accurate (0.9992) vs. Decision Tree (0.6173).


### 6. Produce File of Predictions on Testing Data (n=20)

```{r test_predictions, echo=FALSE}

predRF_testing <- predict(modRF, testing)

answer_file = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("question_#",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

answer_file(predRF_testing)

```

